\bf Model Visualization Techniques for a Social Network Model

### 2017-08-20 paper.Rnw
add a lineup, through new section 4.1.

<<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016. Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25.', fig.height=3>>=
seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
ggplot(data = seobama) + 
  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= F, fiteach = T, linewidth = .25,
           aes(from_id = source, to_id = target, color = party),size = 1) + 
  ThemeNet + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~senate, nrow = 1)
@

@@

@@

In \citeauthor{modelvis}, they define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. What does this definition mean for network data? For dynamic social networks, there are a few different data ``spaces." First, we have the actors and their corresponding covariates. Second, we have the dyad variables that describe the relationships between the variables. Finally, we have the time over which the network is allowed to evolve. These three data pieces can be visualized together in various ways. The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if the dyad between two nodes has value greater than zero, and draws nothing otherwise. The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.

One tool that can bring these three data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}. Different aesthesic aspects of each piece of the node-link diagram can be tied to the underlying node or edge data. The color, size, and shape of the points can be used to represent variables in the node data, and the color, size, and linetype of the links between points can be used to represent the edge variables. In a social network, node data might be age, gender, and occupation of the person in the network, and node data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact. Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.

XXX INSERT AN EXAMPLE OF DATA WITH LOTS OF EDGE, NODE VARS OF INTEREST XXX

These visual connections between the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways. Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix. The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.

Another way to view the model in the data space is through simulation from the model. Since no single simulation from a model is going to look like the data, it's better to visualize many simulations together. For network data, one way to do this is through the traditional node-link diagram. In Figure~\ref{fig:summnetM1}, we show a summary network created with 1,000 simulations of the second wave of the network from Model 1. Only edges that appear in more than 5\% of the simulations (in at least 51 of the 1000 simulations) are included in the visualization, with edges that appear more frequently emphasized by thicker linewidths and a brighter red color. On either side of the summary network, we show the actual data, wave 1 on the left, and wave 2 on the right. \st{We can see that the structure of the summary network is much more similar to the first network observation than to the second network observation. The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}. This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one. I had wave == 2 instead of wave == 1 in the code. This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs. XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure. It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data. This could indicate that we should remove the reciprocity parameter from the model.

@@

\st{Because of the multi-level data structure of a network, viewing the model in the data space can depend on which aspect of the model you are interested in viewing. XXX Ideas: summary statistic dist. on nets simulated from model and show statistic from data; interactive plots showing objective function values on nodes; select node, select other node, show how the function values change interactively (future); goodness of fit should do here. XXX}

>>>>

<<senateplot, fig.cap='The collaboration network in the four senates during the Obama years, 2009-2016. Edges are shown only if the weighted propensity to cosponsor from one senator to another is greater than 0.25. We use the Fruchterman-Reingold algorithm to layout the node-link diagram.', fig.height=3>>=
seobama <- read_csv("data/congress/senateobamapres_gsw_25.csv")
ggplot(data = seobama) + 
  geom_net(directed = T, labelon=F, arrowsize = .25, singletons= F, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party),size = 1) + 
  ThemeNet + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~senate, nrow = 1)
@

@@

XXX TO DO - working on model estimates tonight XXX

@@

The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space. In \citeauthor{modelvis}, they define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. But what does this definition mean for network data? For dynamic social networks, there are a few different data ``spaces": 
\begin{enumerate}
\item the actors and their corresponding covariates,
\item the edges and their correpsonding variables that describe the relationships between the nodes, and 
\item the time over which the network evolves
\end{enumerate}
These three data pieces can be visualized together in various ways. The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise. The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.

One tool that can bring these three data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}. Different aesthesic aspects of each piece of the node-link diagram can be tied to the underlying node or edge data. The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables. In a social network, node data might be age, gender, and occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact, and we can view the network at different timepoints side-by-side to see its evolution. Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.

<<funsenatevis, fig.height=3, fig.cap='The 111th Senate separated into ``BC" and ``AC": before Clinton left the Senate to become Secretary of State (on the right), and after she left (on the left).' >>=
se111clint <- filter(seobama, senate == 111)
se111noclint <- filter(seobama, senate == 111)
se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
se111clint$Clinton <- 'Yes'
se111noclint$Clinton <- 'No'
clintonSenate <- rbind(se111clint, se111noclint)

ggplot(data = clintonSenate) + 
  geom_net(directed = T, labelon=F, arrowsize = .5, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  #theme_net() + 
  ThemeNet + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State. Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, in the middle of her second term in the senate. In that time, she authored two bills and was a cosponsor on 17 other bills. With Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her in the diagram. We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator. We also map shape to sex, and keep color mapped to party of the senator. In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators. 
% ST: I don't like this paragraph
% The visual aspects of the node-link diagram connecting the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways. Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix. The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.

@@

Another way to view the model in the data space is through simulation from the model. No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve. It would therefore better to visualize many simulations together. For network data, one way we propose to do this is through a "summary" or "average" network drawn using the traditional node-link diagram. In Figure~\ref{fig:summnetM1}, we show a summary network created with 1,000 simulations of the second wave of the network from Model 1. \st{To make this summary network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated. We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation. Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations. This is the network we draw using the node-link diagram.} An edge is only drawn in the summary network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color. On either side of the summary network in Figure\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right. \st{We can see that the structure of the summary network is much more similar to the first network observation than to the second network observation. The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}. This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one. I had wave == 2 instead of wave == 1 in the code. This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs. XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure. It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data. This could indicate that we should remove the reciprocity parameter from the model.

<<gof2, fig.cap="An example of viewing the ">>=
qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
  ThemeNoNet + 
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance", subtitle = "Data shown as vertical line")

\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the model and the data you are interested in viewing. The SIENA software includes a method for modelling covariate change in the network by incorporating the network structure, assuming that the ties between nodes can affect how nodes behave over time. In this instance, a time series plot of predicted covariate values for the nodes as the network evolves is viewing the model in the time and node spaces. If, as in Figure~\ref{fig:sienagof}, summary statistics computed on the network are of interest, distributions of these statistics computed on networks simulated from the model show the model in the edge data space. For an example, see Figure~\ref{fig:gof2}, where the distribution of the log of the Mahalanobis distance is shown for the outdegree distribution auxiliary statistics, and the log Mahalanobis distance for the observed data is shown as a vertical line.

\begin{figure}
\includegraphics[width=.75\linewidth]{img/Model1LUSmallFriends.png}
\caption{\label{fig:lineup} A small lineup of node-link diagrams showing the second wave of our small friendship network among five networks simulated from model $M1$.}
\end{figure}

Another potential goodness-of-fit visualization is a lineup like those proposed in \citet{buja:2009}. A \texttt{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p. 4369]{buja:2009}. It can be thought of like a police lineup, where the "suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect. In data and model visualization, the "suspect" is a plot of the true data, while the "filler" is composed of several plots of mock data, simulated from a hypothesized model. If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit. In Figure~\ref{fig:lineup}, the second wave of the small friendship network is shown among five simulated networks from model $M1$ using parameter values estimated from the data. The lineup allows us to view the model in the data space by placing instances of the model side-by-side with the data and examining the differences.}
